Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-06-17 paper/ifo-appendix.tex
working on the viterbi sectoin and further comments from Andrew

To obtain part of the transfer function for the optical microphone, we want to know the resultant change in intensity, $\delta I$, at a point in the interference pattern, given a small change, $\delta d$, in the difference between the arm lengths of the mirrors. Let the point be at angle $\theta$ on the screen (in the far-field) above the central ray as measured from the mirrors. Noting circular symmetry as to not specify exactly where in the fringe the point is.

At this point, two rays meet, one from each mirror, with total electric field amplitude $E_1(t) + E_2(t)$. Given an initial difference between the arm lengths of $d$, the path difference between the two beams at the screen is $2 d \cos{\theta}$ as the longer beam must travel back and forth over the extra distance, and both travel at angle $\theta$ to meet the point~\cite{fringes:online}.

The intensity at the point is then the norm squared of the total electric field amplitude as expressed in Equation~\ref{eq:intensity_derivation} below.

\begin{align}
    E_1(t) &= A e^{i (k L - \omega t)} \\
    E_2(t) &= A e^{i (k (L + 2 d \cos{\theta}) - \omega t)} \\
    I(t) &= \lvert A e^{i (k L - \omega t)} + A e^{i (k (L + 2 d \cos{\theta}) - \omega t)} \rvert^2
\label{eq:intensity_derivation}    
\end{align}

Collapse the phase difference to some value $\phi = 2 k d \cos{\theta}$. Then expand the norm squared using Euler’s relation, $e^{i \theta} = \cos{\theta} + i \sin{\theta}$. Here we stop considering intensity as a function of time as the time dependence will vanish.

\begin{align}    
    I =& A^2 \lvert e^{i (k L - \omega t)} + e^{i (k L + \phi - \omega t)} \rvert^2,\, \phi = 2 k d \cos{\theta} \\
    I =& A^2 ((\cos{(k L - \omega t)} + \cos{(k L + \phi - \omega t)})^2 \nonumber\\ &+ (\sin{(k L - \omega t)} + \sin{(k L + \phi - \omega t)})^2)
\end{align}

Expand and collect like terms, pull out a factor of two, and then recognise the form of $\cos{(\alpha-\beta)}$.

\begin{align}
    I =& A^2 (1 + 2 \cos{(k L - \omega t)} \cos{(k L + \phi - \omega t)} \nonumber\\&+ 1 + 2 \sin{(k L - \omega t)} \sin{(k L + \phi - \omega t)}) \\
    I =& 2 A^2 (1 + \cos{\phi}),\, \phi = \frac{4 \pi d \cos{\theta}}{\lambda}
\end{align}

Finally, changing derivatives and expand $\phi$ to find that the change in intensity given a change in distance is non-linear, that ${\delta I}/{\delta d}$ is non-constant.

\begin{align}    
    \frac{\delta I}{\delta d} &= \frac{\delta I}{\delta \phi}\; \frac{\delta \phi}{\delta d}\\
    \frac{\delta I}{\delta\phi} &= - 2 A^2 \sin{\phi}\\
    \frac{\delta\phi}{\delta d} &= \frac{4 \pi \cos{\theta}}{\lambda}\\
    \therefore \frac{\delta I}{\delta d} &= \frac{- 8 \pi A^2 \cos{\theta}}{\lambda} \sin{(\frac{4 \pi \cos{\theta}}{\lambda} d)}
\end{align}

This non-linearity in part of the transfer function will make the reverse problem of extracting the original signal more difficult.

\section{Implementation of the Viterbi algorithm}
\label{app:viterbi}

We used the Viterbi algorithm to find a signal in a spectrogram. Here we detail the implementation used.

We consider a spectrogram grid where each element in the grid is the Fourier amplitude of a particular frequency at a particular time. These are normalised between $(0, 1)$ to use as multiplicative weights. In fact, we take the logarithm and so the sum of all weights to avoid underflow. The best path has the highest weight among all paths.

With the grid prepared, the algorithm starts in the second column. In each iteration, the algorithm looks for the best path to each cell from the cells in the previous column, selecting the previous cell with the greatest cumulative weight among those ‘nearby’ (the value of which was discovered in the previous iteration). Where the cumulative weight is the weight of the best path to reach that previous cell. The nearness of a previous cell is given by the limit on the allowed frequency change over time.

For each cell in the current column, the algorithm calculates the cumulative weight to reach that cell by adding its weight to the cumulative weight to reach it. The algorithm also notes the row index of which cell in the previous column it best connects to.
Continuing like this until it reaches the end of the grid, the algorithm then looks for the greatest weight in the last column and retraces the indices of the steps it took until it reaches the start of the grid. That discovered best path is the Viterbi path.

@@

>>>>

Part of the transfer function for the optical microphone relates change in intensity, $\delta I$, at a point $P$ in the interference pattern, given a small change, $\delta d$, in the difference between the arm lengths of the mirrors. 
Let $\theta$ be the angle between the optical axis and the ray from the centre of the relevant mirror to the point $P$. 
%Let the point be at angle $\theta$ on the screen (in the far-field) above the central ray as measured from the mirrors. 
%Noting circular symmetry as to not specify exactly where in the fringe the point is.
At $P$, two rays meet, one from each mirror, with total electric field amplitude $E_1(t) + E_2(t)$ and 
\begin{align}
    E_1(t) &= A e^{i (k L - \omega t)},                       \label{eqn:E1} \\
    E_2(t) &= A e^{i (k (L + 2 d \cos{\theta}) - \omega t)}\,  \label{eqn:E2}, 
\end{align}
where $A$ is the amplitude, $L$ is the unperturbed arm length, $k$ is the wavenumber and $\omega$ is the angular frequency of the laser. 
If the unperturbed difference between the arm lengths is $d$, then the path difference between the two rays is $2 d \cos{\theta}$~\cite{fringes:online}.

The intensity $I(t)$ at $P$ is then the norm squared of the total electric field amplitude,
\begin{equation}
I(t) = \lvert A e^{i (k L - \omega t)}  + A e^{i (k (L + 2 d \cos{\theta}) - \omega t)} \rvert^2\,.
\label{eqn:intensity1}    
\end{equation}
Some elementary trigonometry yields
\begin{equation}
%I(t) = 2 A^2 (1 + \cos{\phi})
I(t) = 2 A^2 \left[ 1 + \cos{\left( 2 k d(t) \cos{\theta} \right) } \right]  \label{eqn:intensity2}\,.
\end{equation}
Equation~\ref{eqn:intensity2} shows that $I(t)$ and $d(t)$ are related nonlinearly. 
Near a bright fringe, where one has $\cos{[2kd\cos{(\theta)}]} \approx 1$, the nonlinearity is quadratic, with $\delta I \propto (\delta d)^2$ for small $\delta d$. 
Consequently, an oscillation of the form $\delta d \propto \exp{( -i \Omega t )}$ in the arm length difference produces a DC signal and a frequency-double signal $\propto \exp{(-2i\Omega t)}$ in $\delta I$. 
This is discussed further in Section~\ref{}\han{to do}.

%Collapse the phase difference to some value $\phi = 2 k d \cos{\theta}$. Then expand the norm squared using Euler’s relation, $e^{i \theta} = \cos{\theta} + i \sin{\theta}$. Here we stop considering intensity as a function of time as the time dependence will vanish.

%\begin{align}    
%    I =& A^2 \lvert e^{i (k L - \omega t)} + e^{i (k L + \phi - \omega t)} \rvert^2,\, \phi = 2 k d \cos{\theta} \\
%    I =& A^2 ((\cos{(k L - \omega t)} + \cos{(k L + \phi - \omega t)})^2 \nonumber\\ &+ (\sin{(k L - \omega t)} + \sin{(k L + \phi - \omega t)})^2)
%\end{align}

%Expand and collect like terms, pull out a factor of two, and then recognise the form of $\cos{(\alpha-\beta)}$.

%\begin{align}
%    I =& A^2 (1 + 2 \cos{(k L - \omega t)} \cos{(k L + \phi - \omega t)} \nonumber\\&+ 1 + 2 \sin{(k L - \omega t)} \sin{(k L + \phi - \omega t)}) \\
%    I =& 2 A^2 (1 + \cos{\phi}),\, \phi = \frac{4 \pi d \cos{\theta}}{\lambda}
%\end{align}

%Finally, changing derivatives and expand $\phi$ to find that the change in intensity given a change in distance is non-linear, that ${\delta I}/{\delta d}$ is non-constant.

%\begin{align}    
%    \frac{\delta I}{\delta d} &= \frac{\delta I}{\delta \phi}\; \frac{\delta \phi}{\delta d}\\
%    \frac{\delta I}{\delta\phi} &= - 2 A^2 \sin{\phi}\\
%    \frac{\delta\phi}{\delta d} &= \frac{4 \pi \cos{\theta}}{\lambda}\\
%    \therefore \frac{\delta I}{\delta d} &= \frac{- 8 \pi A^2 \cos{\theta}}{\lambda} \sin{(\frac{4 \pi \cos{\theta}}{\lambda} d)}
%\end{align}

%This non-linearity in part of the transfer function will make the reverse problem of extracting the original signal more difficult.

\section{Viterbi algorithm}
\label{app:viterbi}

This appendix contains some details regarding implementation of the Viterbi algorithm described in Section~\ref{sec:viterbi}. 
The Viterbi algorithm is a classic method in signal processing, whose theoretical underpinnings and implementation are easily accessible to undergraduate students. 
The interested reader is referred to the excellent review paper by Rabiner~\cite{Rabiner:1989} and textbook by Quinn and Hannam~\cite{QuinnEtAl:2001} to learn more.

\han{to do! psudo code here}
%We consider a spectrogram grid where each element in the grid is the Fourier amplitude of a particular frequency at a particular time. These are normalised between $(0, 1)$ to use as multiplicative weights. In fact, we take the logarithm and so the sum of all weights to avoid underflow. The best path has the highest weight among all paths.

%With the grid prepared, the algorithm starts in the second column. In each iteration, the algorithm looks for the best path to each cell from the cells in the previous column, selecting the previous cell with the greatest cumulative weight among those ‘nearby’ (the value of which was discovered in the previous iteration). Where the cumulative weight is the weight of the best path to reach that previous cell. The nearness of a previous cell is given by the limit on the allowed frequency change over time.

%For each cell in the current column, the algorithm calculates the cumulative weight to reach that cell by adding its weight to the cumulative weight to reach it. The algorithm also notes the row index of which cell in the previous column it best connects to.
%Continuing like this until it reaches the end of the grid, the algorithm then looks for the greatest weight in the last column and retraces the indices of the steps it took until it reaches the start of the grid. That discovered best path is the Viterbi path.

@@

%\section{Optical Components List}
%\label{app:opComponents}
%In this appendix we include a parts list for the optical components of the Michelson interferometer. 
%The interferometer described in this paper uses parts from ThorLabs~\cite{}, with a total price of AU\$ 1120 at the time of purchase (July 2019). 
%We also refer the reader to other examples of table-top interferometers at various price points~\cite{TTExhibit:2020,TTExhibit:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999}.

%\begin{table*}
%\begin{tabular}{llll}
%\hline \hline
%Item & Component & Quantity & Part number \\ 
%\hline
%1    & Aluminium optical breadboard $450\,{\rm mm} \times 450\,{\rm mm}$ & 1 & MB4545/M \\
%\hline \hline
%\end{tabular}
%\end{table*}

### 2020-06-17 paper/ifo-complexAudio.tex
working on the viterbi sectoin and further comments from Andrew

We describe additional components to the demonstration in Section~\ref{sec:photodiode} and our initial results in Section~\ref{sec:initialResultsOpMic}.
Filter and speech enhancement techniques applied to the data are detailed in Sections~\ref{sec:simple_filters} and~\ref{sec:logmmse} respectively.
This section is indirectly related to the advanced signal processing required to extract gravitational-wave signals. 
The optical microphone provides an interesting independent demonstration for a broader physics and engineering audience, particularly for the undergraduate lab.

@@

\begin{figure*}
	\includegraphics[width=.6\textwidth]{figures/setup_pic2.pdf}
\caption{Photograph of the optical microphone set-up. The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. The interferometer is laid out as in Figure~\ref{fig:ifo_schematic_podo}. In the main image, the photodiode is placed behind the cloth screen. The inset shows a face-on view of the photodiode with the covering screen removed. See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.}
	\label{fig:setup_pic2}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/filter_timeseries_aa_melatos-cropped.pdf}
	\caption{Time series of the optical microphone recording of an adult male voice (saying ``A cathode ...'') shown before (top panel) and after (bottom panel) the logMMSE estimator is applied. One second of the one minute recording is shown. The rise at the start of the bottom time series is an expected result from filtering a finite signal.}
	\label{fig:logMMSE_timeseries}
\end{figure*}

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/filter_spectrum_aa_melatos-cropped.pdf}
	\caption{Spectrum of optical microphone recording of adult male voice shown before (top panel) and after (bottom panel) the logMMSE estimator is applied. The spectrum shows up to $2\,{\rm kHz}$ and to amplitude $2$ to see greater detail.}
	\label{fig:logMMSE_spectrum}
\end{figure*}

>>>>

We describe additional components to the demonstration in Section~\ref{sec:photodiode} and initial results in Section~\ref{sec:initialResultsOpMic}.
Filter and speech enhancement techniques applied to the data are detailed in Sections~\ref{sec:simple_filters} and~\ref{sec:logmmse} respectively.
This section is indirectly related to the advanced signal processing required to extract gravitational-wave signals. 
The optical microphone provides an interesting independent demonstration for a broader physics and engineering audience, particularly for the undergraduate lab.

@@

\begin{figure*}
	\includegraphics[width=.6\textwidth]{figures/setup_pic2.pdf}
\caption{Photograph of the optical microphone. The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. The interferometer is laid out as in Figure~\ref{fig:ifo_schematic_podo}. In the main image, the photodiode is placed behind the cloth screen. The inset shows a face-on view of the photodiode with the covering screen removed. See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.}
	\label{fig:setup_pic2}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/filter_timeseries_aa_melatos-cropped.pdf}
	\caption{Time series of the optical microphone recording of an adult male voice (saying ``A cathode ...'') shown before (top panel) and after (bottom panel) the logMMSE estimator is applied. One second of the one minute recording is shown. The rise at the start of the bottom panel is expected when filtering a finite signal.}
	\label{fig:logMMSE_timeseries}
\end{figure*}

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/filter_spectrum_aa_melatos-cropped.pdf}
	\caption{Spectrum of optical microphone recording of the same adult male voice in Fig.~\ref{fig:logMMSE_timeseries} shown before (top panel) and after (bottom panel) the logMMSE estimator is applied. The spectrum shows up to $2\,{\rm kHz}$ and to amplitude $2$ to see greater detail.}
	\label{fig:logMMSE_spectrum}
\end{figure*}

### 2020-06-17 paper/ifo-conclusions.tex
working on the viterbi sectoin and further comments from Andrew

In Sections~\ref{sec:ifo} and ~\ref{sec:single_tone} we demonstrate the recovery of signals with constant and slowly wandering frequency. 
A webcam sampling at \han{XX}$\,{\rm Hz}$ is used to track the interference pattern and the injected signal is recovered succesfully with the Viterbi algorithm. 
These sections direcly demonstrate the continuous wave search techniques used in gravitational-wave data analysis. 
%Using a webcam to track the interference pattern, we inject a wandering frequency signal into the system with a speaker attached to the back of one of mirrors of the interferometer. We then demonstrate how to recover this signal using the Viterbi algorithm, just as is done in continuous gravitational wave searches.

@@

The demonstration presented here has good potential to be adapted for use in the undergraduate laboratory. 
It can be used to demonstrate topics of interest to physicists and electrical engineers including: interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement. 
%These demonstrations have good potential to be adapted into the undergraduate laboratory. There is much to explore in the physics of the interferometer and the astronomy of the motivating continuous gravitational wave searches. Most of all, the signal processing of speech enhancement is an interesting problem that’s relevant to digital communications.

>>>>

In Sections~\ref{sec:ifo} and ~\ref{sec:single_tone} we demonstrate the recovery of signals with constant and slowly wandering frequency. 
A webcam sampling at $30\,{\rm Hz}$ is used to track the interference pattern and the injected signal is recovered succesfully with the Viterbi algorithm. 
These sections direcly demonstrate the continuous wave search techniques used in gravitational-wave data analysis. 
%Using a webcam to track the interference pattern, we inject a wandering frequency signal into the system with a speaker attached to the back of one of mirrors of the interferometer. We then demonstrate how to recover this signal using the Viterbi algorithm, just as is done in continuous gravitational wave searches.

@@

The demonstration presented here has good potential to be adapted for use in the undergraduate laboratory. 
It can be used to demonstrate topics of interest to physicists and electrical engineers including: interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement. 
The detection of gravitational waves in 2015 led to increased excitement and public interest in gravitational wave science. 
The table-top demonstration presented here can also be adapted as a tool to explain gravitational wave research developments for a wider non-specialist audience. 
The future will bring many more detections, including further observations of transient signals from binary black hole and binary neutron star mergers, as well as the anticipated first detections of other classes of signal, including continuous waves.

%These demonstrations have good potential to be adapted into the undergraduate laboratory. There is much to explore in the physics of the interferometer and the astronomy of the motivating continuous gravitational wave searches. Most of all, the signal processing of speech enhancement is an interesting problem that’s relevant to digital communications.

### 2020-06-17 paper/ifo-futureWork.tex
working on the viterbi sectoin and further comments from Andrew

% wild speculation
We are currently working towards two future advancements for the optical microphone.

The first is to better isolate the experiment from the mains supply. 
The Raspberry Pi is USB powered and requires low power so it could be run off a commercial external phone battery. 
Everything else in the circuit, including the photodiode, is connected to the Pi, so this would isolate them electrically as well. 
The laser is DC powered through a converter, so should not introduce mains harmonics itself. 
Using the interferometer in a dark room and a sealed container could better isolate if from ambient light and air currents.
These changes may reduce or remove the $50\,{\rm Hz}$ harmonics from the data.
%With the interferometer in a dark room and under a box to better isolate from any ambient light or air currents, this would hopefully remove (or better separate) all sources of the 50Hz mains hum from the optical microphone.

The second advancement would be to achieve a full model of the transfer function of the system. 
Theoretically, if we had the correct transfer function, then signal recovery would only be limited by the signal-to-noise ratio as an optimal matched filter could be constructed.
The transfer function would need to start from the voltage sent to the speaker and end with the value recorded by the Pi, and include any non-linearities in the speaker audio production, the speaker-mirror coupling, the separation-intensity relation, and the photodiode reading. 
The sequence that the signal takes is shown in Figure~\ref{fig:pipeline_highlighted}.
In our investigation, we have currently only derived part of this sequence, the relation between a deflection of the mirror and the change in the intensity of the interference pattern at some point. 
We find this relation to be non-linear, see Appendix~\ref{app:intensity_derivation} for the derivation.

Asides from the transfer function, for extracting the speech there exists much literature~\cite{HMM_english} on using a hidden Markov model train to English phonemes to recognise speech.
Alternatively, machine learning solutions exist throughout the field that are able to compete~\cite{SEGAN} with statistical techniques like the logMMSE estimator.

To remove the mains hum we could also monitor it separately from the interferometer and subtract the sinusoids with the right phase and amplitude in real (not frequency) space to avoid the problems of Section~\ref{sec:simple_filters}. 
This could be done with the existing signals if the recording duration is long enough to precisely measure the phase and amplitude over a large number of cycles.

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/pipeline_highlighted.pdf}
	\caption{Sequence of relations for optical microphone transfer function. The currently derived relation is highlighted.}
	\label{fig:pipeline_highlighted}
\end{figure}

>>>>

% wild speculation
We are currently working towards two improvements of the optical microphone.

The first is to better isolate the experiment from the mains supply. 
The Raspberry Pi is USB powered and requires low power so it could be run off a commercial external phone battery. 
Everything else in the circuit, including the photodiode, is connected to the Pi. 
The laser is DC powered through a converter, so should not introduce mains harmonics itself. 
%With the interferometer in a dark room and under a box to better isolate from any ambient light or air currents, this would hopefully remove (or better separate) all sources of the 50Hz mains hum from the optical microphone.

The second improvement is to measure the full model of the transfer function of the system. 
Theoretically, with the correct transfer function, signal recovery is limited by the signal-to-noise ratio as an optimal matched filter could be constructed.
The transfer function starts from the voltage sent to the speaker and end with the value recorded by the Pi, and include any non-linearities in the speaker audio production, the speaker-mirror coupling, the separation-intensity relation Appendix~\ref{app:intensity_derivation}, and the photodiode reading. 
A schematic of the full optical microphone transfer function is shown in Figure~\ref{fig:pipeline_highlighted}.

A voluminous literature exists about using a hidden Markov model trained against phonemes to recognize speech~\cite{HMM_english}. 
Interestingly, this connects back to the Viterbi algorithm in Section IV, which is also underpinned by a hidden Markov model, albeit of a different type. 
Alternatively, machine learning solutions exist throughout the field that are able to compete~\cite{SEGAN} with statistical techniques like the logMMSE estimator.

%To remove the mains hum we could also monitor it separately from the interferometer and subtract the sinusoids with the right phase and amplitude in real (not frequency) space to avoid the problems of Section~\ref{sec:simple_filters}. 
%This could be done with the existing signals if the recording duration is long enough to precisely measure the phase and amplitude over a large number of cycles.

\begin{figure*}
	\includegraphics[width=.99\textwidth]{figures/pipeline_highlighted.pdf}
	\caption{
Flowchart for the optical microphone transfer function. 
The highlighted stage is analysised in Appendix XX, the other stages are left to future work. 
}
	\label{fig:pipeline_highlighted}
\end{figure*}

### 2020-06-17 paper/ifo-wanderingTone.tex
working on the viterbi sectoin and further comments from Andrew

\begin{figure}
\includegraphics[width=0.49\textwidth]{/home/hannahm/Documents/viterbi/quickTest/circlesTest}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from $t_0$ on the left and ending at $t_4$ on the right. 
The color of the box indicates the likelihood that a signal is present in the box, where a lighter color corresponds to higher likelihood. 
Each box has a single path leading to it indicated by the dotted lines. 
The Viterbi path is highlighted in orange. 
At each time step, the paths can move up one $f$ box, move down one $f$ box, or stay in the same $f$ box. 
See the main text for a full description. 
}
\end{figure}

@@

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. $Pr[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
%If there is equal probability of starting $f_j$ at $t=t_0$, the probability of path $Q$ given the observation $O$ is 
\begin{eqnarray}
%P(Q|O) =& A(f_j) F(t=N_t,f_j), \nonumber \\ &~~ \times \dots, A(f_j) F (t=t_0,f_j).
Pr[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &*A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     & *Pr[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes $Pr[f(t_{N_t})] = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm is an efficient method to find the Viterbi path. 
Here we describe the algorithm as psudeo-code referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements of the spectrogram grid (lighter colours correspond to higher likelihood). 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at $t_1$, each $f_j$ element has three possible paths it could originate from at $t_0$, indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ element, the highest $F[t_0,f_j] A[f(t_0),f(t_1)]$ from the previous time step is selected, shown by the solid lines in Fig.~\ref{fig:viterbi}. 
For example for the element labelled (a), the highest value is directly behind it, therefore the transition from $\{t_0,f_0\}$ to $\{t_1,f_0\}$ is selected. 
Each element has now selected the best path from the previous time step. 
In Fig.~\ref{fig:viterbi}, black lines show the best paths and grey lines show all other paths. 
\item Moving to $t_2$, the path with the highest value of the recursive relation in equation~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}. 
\item Steps 2 is repeated until the end of the grid is reached. 
\item At $t=t_{N_t}$, each $f_j$ has a value $Pr[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The highest likelihood path for each $f_j$ can be seen by tracing the solid lines from right to left in Fig.~\ref{fig:viterbi}. 
The $f_j$ with the maximum $Pr[f(t_{N_t})=f_j]$ is labelled (b) in Fig.~\ref{fig:viterbi}. 
\item Tracing the p
At the end of the grid, each $f_j$ has a unique path associated with it (seen by tracing the dotted lines from right to left in Fig.~\ref{fig:viterbi}). The element with the highest accumulated probability is selected, in the case of Fig.~\ref{fig:viterbi}, this is labelled (b). 
\item Tracing the path from right to left in Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). This is the Viterbi path, which is highlighted in orange in Fig.~\ref{fig:viterbi}. 
\end{enumerate}

>>>>

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The color of each circle indicates the likelihood that a signal is present in the box, where a lighter color corresponds to higher likelihood. 
AT each time step, the paths can move up one $f$ step, move down one $f$ step, or stay at the same $f$ step (shown by lines in the diagram). 
Each circles has three paths leading to it indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by equation~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. $Pr[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
%If there is equal probability of starting $f_j$ at $t=t_0$, the probability of path $Q$ given the observation $O$ is 
\begin{eqnarray}
%P(Q|O) =& A(f_j) F(t=N_t,f_j), \nonumber \\ &~~ \times \dots, A(f_j) F (t=t_0,f_j).
Pr[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &*A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     & *Pr[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes $Pr[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm is an efficient method to find the Viterbi path. 
Here we describe the algorithm as psudeo-code referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements of the spectrogram grid (lighter colours correspond to higher likelihood). 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at $t_1$, each $f_j$ element has three possible paths it could originate from at $t_0$, indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ element, the highest $F[t_0,f_j] A[f(t_0),f(t_1)]$ from the previous time step is selected, highlighted by the black lines in Fig.~\ref{fig:viterbi}. 
For example for the element labelled (a), the highest value is directly behind it, therefore the transition from $\{t_0,f_0\}$ to $\{t_1,f_0\}$ is selected. 
Each element has now selected the best path from the previous time step. 
In Fig.~\ref{fig:viterbi}, black lines show the best paths and grey lines show all other paths. 
\item Moving to $t_2$, the path with the highest value of the recursive relation in equation~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}. 
\item Steps 2 is repeated until the end of the grid ($t=t_{N_t}$) is reached. 
Only the best paths are stored. 
\item At $t=t_{N_t}$, each $f_j$ has a value $Pr[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.
\item The Viterbi path is then found by backtracking along the best paths at each time step. 
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighed in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
It is also very computationally efficient.
